Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2018-09-27 astronum_2018/astronum-xrb.tex
more work on the algorithm section

We have begun a set of flame spreading calculations using the
compressible hydrodynamics code Castro~\cite{castro}, part of the
AMReX astrophysics suite~\cite{astronum:2017}.  \castro\ uses adaptive
mesh refinement, supports a general equation of state and reaction
networks, has a conservative gravity formulation~\cite{wdmergerI}, and
is optimized to run on current supercomputers.  All of our code base
is open source and available on
github\footnote{\url{http://github.com/amrex-astro/}}, and all of the
solvers, problem setup, initial models, and analysis scripts for each
of our science problems are distributed with the codes.

@@

Machine architectures are evolving, with heterogeneous architectures
becoming more common.  Our recent development in \castro\ has focused
on moving the hydrodynamics to GPUs~\cite{astronum:2017}.  The XRB
problem is an ideal case for this, since hydrodynamics with constant
gravity (for plane-parallel domains) ports in a straightforward manner
to GPUs.  The current GPU version of the \castro hydrodynamics solver
is about $10\times$ faster using GPUs on a node than the entire node
of CPU cores.  We see similar speed ups with reaction networks, with
these GPU ports under active development.  This means a GPU
calculation could lower the above costs by a factor of 10.  GPU
offloading of the reaction networks will also allow us to explore
larger networks and more detailed nuclear physics.

If we continue to resolve the flame structure, then we need algorithms
that can do so less expensively.  We can be smarter about our grid
refinement strategy, putting the finest grids only at the very base of
the burning layer, where the flame is the thinnest.  Using
well-balanced schemes~\cite{kappeli:2016} can help us retain
hydrostatic equilibrium (HSE) in the lower refined regions, reducing
the number of zones needed to maintain fidelity.  We can also explore
initializing newly created fine grids ahead of the burning with an
understanding of hydrostatic equilibrium to allow us to only refine
the peak burning as the flame moves across the star.

Another way we can meet the resolution requirements is to use a more
accurate hydrodynamic method.  In particular, we are developing fully
fourth-order (in space and time) coupling of hydrodynamics and
reactions using the method of spectral deferred corrections (SDC) in
\castro, instead of the more commonly employed Strang splitting.  This
provides us with two benefits.  First, the improved coupling actually
reduces the stiffness of the reactions, requiring fewer righthand side
calls and therefore reducing the expense of the
reactions. \MarginPar{need some refs here} Second, by moving to fourth
order, we may also be able to reduce our resolution requirements
needed for converged flames, and since computational work scales like
$(\Delta x)^4$ in 3D, this can save a lot of computer time.

As an example of the improvements in coupling between hydro and \MarginPar{Don can add some comments about startup costs?}
reactions, Figure~\ref{fig:sdc} shows the mass fraction of helium
behind a detonation over the course of 3 timesteps.  The points
represent individual calls to the righthand side function of our
reaction network (a 13-isotope $\alpha$-chain), and the density of
points indicates how hard the integrator is working.  We use the same
tolerances for both the Strang-split method and the SDC method.  For
the SDC method, we are using a second-order accurate method.  We
predict a time-centered advection term, $\Adv{\Uc}^{n+1/2}_i$ using
standard unsplit Godunov methods, but explicitly include a reactive
source term in the tracing of the interface states, ${\bf R}(\Uc_i)$.  We can then solve
the reactive system, using this advective term as a
piecewise-constant-in-time source
\begin{equation}
\label{eq:syst}
\frac{d{\Uc_i}}{dt} = -\Adv{\Uc}^{n+1/2}_i + {\bf R}(\Uc_i)
\end{equation}
This system can be integrated with standard ODE methods.  To achieve
second-order accuracy, we need to iterate, using the updated $\Uc$ to
create a reactive source included in the predictor for the advective
update, and then reintegrate the system.  The result of this iteration
is that the advection sees the effects of reactions over the timestep
and the reactions see the effects of advection.  For the zone tracked
in Figure~\ref{fig:sdc}, each SDC iteration called the righthand side
function of the network 5 times less than the Strang case.
We are continuing to develop this new integration strategy, with full
fourth-order in space and time reacting hydrodynamics methods to
follow shortly.

@@

Within the realm of compressible hydrodynamics, there are a number of
other techniques that should be explored.  smarter refinement with AMR
+ hydrostatic refinement well-balanced method to allow ultralow
resolution to slide along with the burning front higher-order This are
all

Other ways to make the simulation less expensive can include a subgrid
model, Flame models are commonly used for Type Ia supernovae.  There
however, the flame is thin compared to the scale height, so a simple
model can represent the unresolved flame on the grid.  For the XRB,
the flame thickness is a non-trivial fraction of the pressure scale
height and further, the laminar flame properties will change strongly
over the height of the burning layer (see~\cite{Timmes00}).  This
makes traditional flame models a poor fit.  Methods that use even
something as simple as subzone burning (e.g.~\cite{Wang2012190}).  This is harder...

Finally, we can try hybrid model approaches where the fully
compressible simulation can act as a subgrid model.  \MarginPar{this ties into Ann's stuff, and also the stuff Alice was doing}
Multiscale methods can combine different hydro models together to allow 
for a full star simulation with resolved burning.  Examples include...

>>>>

We have begun a set of flame spreading calculations using the
compressible hydrodynamics code \castro~\cite{castro}, part of the
AMReX astrophysics suite~\cite{astronum:2017}.  \castro\ uses adaptive
mesh refinement, supports a general equation of state and reaction
networks, has a conservative gravity formulation~\cite{wdmergerI}, and
is optimized to run on current supercomputers.  All of our code base
is open source and available on
github\footnote{\url{http://github.com/amrex-astro/}}, and all of the
solvers, problem setup, initial models, and analysis scripts for each
of our science problems are distributed with the codes.

@@

Machine architectures are evolving, with heterogeneous architectures
becoming more common.  Our recent development in \castro\ has focused
on moving the hydrodynamics to GPUs~\cite{astronum:2017}.  The XRB
problem is an ideal case for this, since hydrodynamics with constant
gravity (for plane-parallel domains) ports in a straightforward manner
to GPUs.  The current GPU version of the \castro\ hydrodynamics solver
is about $10\times$ faster using GPUs on a node than the entire node
of CPU cores.  We see similar speed ups with reaction networks, with
these GPU ports under active development.  This means a GPU
calculation could lower the above costs by a factor of 10.  GPU
offloading of the reaction networks will also allow us to explore
larger networks and more detailed nuclear physics.

Another way we can meet the resolution requirements is to use a more
accurate hydrodynamic method.  We are developing fully
fourth-order (in space and time) coupling of hydrodynamics and
reactions using the method of spectral deferred corrections (SDC) in
\castro, instead of the more commonly employed Strang splitting.  This
provides us with two benefits.  First, the improved coupling actually
reduces the stiffness of the reactions, requiring fewer righthand side
calls and therefore reducing the expense of the
reactions. \MarginPar{need some refs here} Second, by moving to fourth
order, we may also be able to reduce our resolution requirements
needed for converged flames, and since computational work scales like
$(\Delta x)^4$ in 3D, this can save a lot of computer time.

As an example of the improvements in coupling between hydro
and \MarginPar{Don can add some comments about startup costs?}
reactions, Figure~\ref{fig:sdc} shows the mass fraction of helium
behind a detonation over the course of 3 timesteps.  The points
represent individual calls to the righthand side function of our
reaction network (a 13-isotope $\alpha$-chain), and the density of
points indicates how hard the integrator is working.  We use the same
tolerances for both the Strang-split method and the SDC method.  For
the SDC method, we are using a second-order accurate method based on
\cite{SDC-old}.  We predict a time-centered advection term,
$\Adv{\Uc}^{n+1/2}_i$ using standard unsplit Godunov methods, but
explicitly include a reactive source term in the tracing of the
interface states, ${\bf R}(\Uc_i)$.  We can then solve the reactive
system, using this advective term as a piecewise-constant-in-time
source
\begin{equation}
\label{eq:syst}
\frac{d{\Uc_i}}{dt} = -\Adv{\Uc}^{n+1/2}_i + {\bf R}(\Uc_i)
\end{equation}
This system can be integrated with standard ODE methods.  To achieve
second-order accuracy, we need to iterate, using the updated $\Uc$ to
create a reactive source included in the predictor for the advective
update, and then reintegrate the system.  The result of this iteration
is that the advection sees the effects of reactions over the timestep
and the reactions see the effects of advection.  For the zone tracked
in Figure~\ref{fig:sdc}, each SDC iteration called the righthand side
function of the network 5 times less than the Strang case.
We are continuing to develop this new integration strategy, with full
fourth-order in space and time reacting hydrodynamics methods to
follow shortly.  The XRB simulations are our target application.

@@

Other ways to make the simulation less expensive can include a subgrid
model, Flame models are commonly used for Type Ia supernovae.  There
however, the flame is thin compared to the scale height, so a simple
model can represent the unresolved flame on the grid.  For the XRB,
the flame thickness is a non-trivial fraction of the pressure scale
height and further, the laminar flame properties will change strongly
over the height of the burning layer (see~\cite{Timmes00}).  This
makes traditional flame models a poor fit.  Methods that use even
something as simple as subzone burning (e.g.~\cite{Wang2012190}) may
help capture the energetics with lower resolution.  This can be
implemented using the AMR machinery.  This may help a little, but will
still be insufficient for full star models.

If we continue to resolve the flame structure, then we need to be
smarter (and more aggressive) about our grid refinement strategy.
Currently, we put the entire fuel layer on the finest grid, but for
the material ahead of the flame, we could keep it at lower resolution
until the flame reaches it.  The main obstacle is that we need to keep
the fuel ahead of the flame in hydrostatic equilibrium, and keeping it
refined accomplishes this well.  Using well-balanced
schemes~\cite{kappeli:2016} can help us retain hydrostatic equilibrium
(HSE) in the lower refined regions, reducing the number of zones
needed to maintain fidelity, but we would still need a way to regrid
the layer ahead of the flame to higher resolution to accurately
capture the burning dynamics.  To accomplish this, we envision
building a refinment strategy that does interpolation of the data onto
the newly refined grids enforcing hydrostatic equilibrium.  This can
allow us to keep only a narrow fully refined region that ``slides''
along the surface of the neutron star keeping pace with the advancing
flame.  Such a scheme would greatly reduce the computational costs.
This is a longer term development.

All of the above developments still involve fully compressible
hydrodynamics, but multiscale methods where we couple different hydrodynamics
solvers together in a single simulation may ultimately be needed to model the full star while resolving the burning front. \MarginPar{Alice and Ian contribute here}

%% Finally, we can try hybrid model approaches where the fully
%% compressible simulation can act as a subgrid model.  
%% Multiscale methods can combine different hydro models together to allow 
%% for a full star simulation with resolved burning.  Examples include...
